Spontaneous breaking of multipole symmetries

### 2021-12-10 multipoleSSB/charlies_edits.tex
charlies edits

Reference~\cite{Gromov2019} includes discussion of some multipole groups; we will review a few here. The simplest case is the maximal multipole group $\mathcal{M}^a_\text{max}$, which includes all shifts by polynomials of degree $a$ or less. Individual polynomials can be written as
\begin{align}
P_c^{I_c} = \mu^{I_c}_{i_1\dots i_c}x^{i_1}\dots x^{i_c}, \label{eqn:basis}
\end{align}
where each  matrix  $\mu^{I_c}_{i_1\dots i_c}$ is fully symmetric and $c\le a$. This group also includes all translations and rotations.

An example of a multipole group that contains all translations and rotations but is \emph{not} the maximal multipole group is the group generated by shifts of the form
\begin{align}
\phi \goesto \phi + \lambda_0 P_0^0 + \lambda_{i} P^{i}_1 + \lambda_{I_2} P^{I_2}_2,
\end{align}
where the degree-0 polynomial is $P_0^0=1$. The other polynomials are
\begin{align}
P_1^{i} = x^i,\quad \quad P_2^{I_2} &= \mu^{I_2}_{ij} x^i x^j,
\end{align}
where $\mu^{I_2}$ are a basis for the rank-2 traceless symmetric $d\times d$ matrices. Let us call this group $\mathcal{M}^2_{\text{sym}}$. Recall that the maximal quadrupole group $\mathcal{M}^2_{\text{max}}$ is already only built from symmetric matrices $\mu$. The tracelessness condition thus only removes one polynomial from the set.

@@

We can now consider building IR field theories for phases in which a multipolar symmetry is completely spontaneously broken, with no residual unbroken subgroup. Reference~\cite{Gromov2019} describes the process in detail, so in the following we will be somewhat succinct. 
We will continue to write things down in terms of the compact scalar $\phi$, which (at in the cases for which SSB is not preempted by strong fluctuations) is to be viewed as the Goldstone for the sponatneously broken symmetry. 
% References~\cite{Griffin2015, Gromov2019} discuss quantum field theories where configurations are weighted by an action. We will also want to consider thermal field theories with Hamiltonian descriptions, so we will abstractly discuss ``kinetic terms" in this section.

Since we are only interested in the low-energy physics of the putative symmetry-broken phase, constructing an appropriate IR field theory amounts to nothing more than constructing an appropriate kinetic term for $\phi$. 
To find a kinetic term invariant under a given multipole group $\mathcal{M}$, we need to find operators $D$ built out of spatial derivatives that annihilate all the polynomials in $\mathcal{M}$. If $a_\text{max}$ is the highest degree of the polynomials in $\mathcal{M}$, the simplest derivative operators which generically do the job are of the form \ethan{may or may not want to switch to the composite index notation $q^{I_{a_{\rm max}+1}} \p_{I_{a_{\rm max}+1}}$ employed later}
\begin{align}
D = q^{i_1\dots i_{a_\text{max}+1}}\partial_{i_1}\dots \partial_{i_{a_\text{max}+1}},
\end{align}
for some symmetric tensor $q^{i_1\dots i_{a_\text{max}+1}}$. Although it is not generically possible to do so~\cite{Gromov2019}, we can sometimes find a set of $D_\alpha$ ($\alpha$ is an abstract index) with $s\le a_{\text{max}}$, where $a_\text{max}$ is the highest degree of the polynomials. In this case, the effective field theory will be invariant under some non-maximal multipole symmetry.
% We say that $D$ is order-$s$ if the maximum number of derivatives appearing in $D$ is $s$.

Writing the invariant derivative operators as $D_\alpha$, the most general kinetic term is ~\cite{Gromov2019} 
\begin{align}
K[\phi(x)] &= g_{\alpha\beta} (D_\alpha \phi) (D_\beta \phi)
\end{align}
for some symmetric tensor $g_{\alpha\beta}$.
We will often write the Fourier transform of the kinetic term as $\phi_{-k} K_k \phi_k$.  Requiring some spatial symmetries restricts the choices of $g_{\alpha \beta}$. For the maximal multipole group, enforcing all rotation symmetries results in the kinetic term
\begin{align}
K_a[\phi(x)] &= (\partial_{i_1}\dots \partial_{i_{a+1}} \phi) (\partial_{i_1} \dots \partial_{i_{a+1}} \phi),
\end{align}
which is the kinetic terms studied in Ref.~\cite{Griffin2015}.

@@

Consider first a clean classical system at $T>0$ with a spontaneously broken maximal multipole symmetry of degree $a$.
Then the kinetic term is proportional to 
\begin{align}
K[\phi(x)] = (\partial_{i_1} \dots \partial_{i_{a+1}} \phi)^2. \label{eqn:adisp}
\end{align} 
Now consider nucleating domains of linear size $L$ in which the expectation value of the order parameter --- namely $\langle e^{i\phi}\rangle$ --- varies by an amount of order 1. Since the order parameter is continuous there will be a thick domain ``wall" over which the order parameter changes gradually from its value in the background phase to its value in the domain. The thickness will be of order $L$.
Equation~\ref{eqn:adisp} tells us the  energy cost of the field changing its value will be minimized if it varies as a polynomial with degree $a+1$. In that case the energy density will scale as $K_k \sim k^{2(a+1)} \sim L^{-2(a+1)}$. Integrating that energy density over a region of size $L$ gives a total energy cost to nucleating domains of $L^{d-2(a+1)}$.

@@

We can also consider an anisotropic multipole group. 
Let the conserved charges be the monopole moment and $d_2$ components of the dipole moment. Then the dispersion will be $k^4 + p^2$, where $k$ has $d_2$ components, $p$ has $d_1$ components, and $d = d_1+d_2$. Classically, the correlation function is
\begin{align}
C &\sim \int \frac{d^{d_2} k \, d^{d_1} p}{k^4 + p^2}\nn
&\sim \int x^{d_2/2 -1} p^{d_1-1} \frac{dx\, dp}{x^2 +p^2} \nn
&\sim \int q^{d_2/2 + d_1 - 2}\frac{dq}{q},
\end{align}
where $q^2 = x^2 + p^2 = k^4 + p^2$. 
If $d_1=2$ the symmetry can be broken for any $d_2>0$, and if $d_1=1$ then $d_2=2$  is critical. Of course, if $d_1=0$ we have an isotropic quartic dispersion and $d=d_2=4$ is critical.

@@

The quantum correlation function behaves as
\begin{align}
C &\sim \int \frac{d\omega d^{d_2} k \, d^{d_1} p}{\omega^2 + k^4 + p^2}\nn
&\sim \int k^{d_2 - 2} p^{ (d_1-1)} \frac{k\,dk\, dp}{\sqrt{k^4 + p^2}} \nn
&\sim \int x^{d_2/2-1} p^{d_1-1} \frac{dx\, dp}{\sqrt{x^2 + p^2}} \nn
&\sim \int q^{d_2/2 -1} q^{d_1 -1} dq,
\end{align}
where again $q^2 = x^2 + p^2 = k^4 + p^2$. This diverges when $d_2/2 + d_1 \le 1$.

The previous analysis suggests a procedure for finding the critical dimension in clean anisotropic systems for breaking the multipole symmetry to the trivial group. First, sort each dimension by the degree of its dispersion relation, so that there are $d_n$ dimensions with dispersion $k^{2n}$. Then define the effective dimension as 
\begin{align}
d_\eff = \sum_n \frac{d_n}{n}.
\end{align}
In a classical system, we then conclude that SSB cannot occur if $d_\eff\le 2$, while in a quantum system the critical dimension is $d_\eff = 1$.

@@

Let us define the compressibilities $\kappa_c$ by 
\be \label{kappadef} \kappa_c^{I_c} = \frac{d\langle n_c^{I_c}\rangle}{d\mu_c^{I_c}},\ee 
where $n_c$ is the number density of $c$-pole charges, $\mu_c$ is a conjugate chemical potential, and $I_c = \{i_1,\dots,i_c\}$ is a composite index.

Let $b\leq a$ denote the smallest degree of multipole charge with nonzero compressibility, i.e. consider a situation in which $\kappa_d>0$ for all $b\leq d\leq a$, while $\kappa_d=0$ for all $d<b$. Then the minimal IR Lagrangian for the symmetry broken phase looks schematically like 
\be \label{generall} \mathcal{L} = \kappa_b (\partial_\tau \phi_{I_b})^2 + g_{J_{a-b+1}}(\partial_{J_{a-b+1}}\phi_{I_b})^2,\ee 
where 
%$I_b = \{i_1,\dots,i_b\}$ and $J_{a-b+1}=\{j_1,\dots,j_{a-b+1}\}$ are composite indices with 
$\p_{J_{a-b+1}} \equiv \p_{j_1} \cdots \p_{j_{a-b+1}}$, and where $\phi_{I_b}$ is a field that shifts by a polynomial of degree $a-b$ under the action of $\mmax a$. The compressibility appears as the coefficient of the time derivative term since $\p_\tau \phi_{I_b}$ is the momentum conjugate to $\phi_{I_b}$ in the quantum theory, and hence correlation functions of fluctuations in $n_b$ are determined by those of $\p_\tau \phi_{I_b}$.

@@

However, a nonzero $\kappa_b$ does not {\it necessarily} imply that $\mmax b$ is spontaneously broken. $\kappa_b\neq 0$ means that the system possesses gapless modes created by $n_b$, but it is possible for fluctuations to be strong enough so that these modes cannot be associated with the Goldstone modes of a spontaneously broken $\mmax b$. Thus the theory \eqref{generall} does not necessarily describe a phase where $\mmax b$ is spontaneously broken, and the exact pattern of symmetry breaking depends on the spatial dimension and whether or not $T$ is nonzero. For example, we can again consider the theory in \eqref{singlescalar} at $d=3,T>0$: even though the charge compressibility $\kappa_0$ (the coefficient of $(\p_\tau\phi)^2$ in \eqref{singlescalar}) is nonzero, $\phi$ fluctuates strongly enough that the monopole symmetry is not spontaneously broken.

@@

%In a classical system at $T>0$ the correlation functions of this operator are
%\begin{align}
%\left\langle \partial_i\phi(x) \partial_j\phi(0) \right\rangle &\sim \delta_{ij} \int \frac{d^dk\, k^2}{k^4},
%\end{align}
%so that the critical dimension is $d=2$, as in the standard Mermin-Wagner theorem for monopole symmetries. This allows us to interpret the dipole charges of the monopole symmetry as monopole charges of the dipole part of the symmetry group.
%However, this interpretation does not hold for  quantum systems. 
%The correlation function scales as
%\begin{align}
%\left\langle \partial_i\phi(x) \partial_j\phi(0) \right\rangle &\sim \delta_{ij} \int \frac{d\omega\, d^dk\, k^2}{\omega^2 + k^4}\nn
%&= \delta_{ij} \int \frac{d^dk\, k^2}{k^2},
%\end{align}
%which is always IR-finite. We thus conclude that at $T=0$, $\mcm^1_{\rm max}$ can be spontaneously broken down to $\mcm^0$ in any dimension

%We then recognize that these fundamental dipoles, which are of course the dipole charges of the monopole symmetry, are also the monopole charges of the dipole symmetry.  We should then expect symmetry breaking cannot occur for $d\le 2$ classically or $d\le 1$ quantum mechanically.
%
%
%In more generality, for $G = \mathcal{M}^{a}_\text{max}$ and $H = \mathcal{M}^b_\text{max}$ there will be a charged operator $\partial_{i_1} \dots \partial_{i_{b+1}}\phi$ with a dispersion as in Eqn.~\ref{eqn:adisp}. Because $G$ is a maximal multipole group, we can still write the polynomials as in Eqn.~\ref{eqn:basis}. Recall that $P_c^{I_c}$ is a degree-$c$ monomial. As long as $\phi$ transforms as 
%\begin{align}
%\delta \phi = \lambda_{I_c} P_c^{I_c}, \label{eqn:partial}
%\end{align}
%then $\partial_{i_1} \dots \partial_{i_{b+1}} \phi$ is invariant under $H=\mathcal{M}^b_\max$. 
%
%Classical correlation functions for these operators take the form
%\begin{align}
%C &= \left\langle \partial_{i_1} \dots \partial_{i_{b+1}} \phi(x) \partial_{j_1} \dots \partial_{j_{b+1}} \phi(0) \right\rangle \nn
%&\sim \int \frac{d^dk\, k^{2(b+1)}}{k^{2(a+1)}},
%\end{align}
%and diverge when $d \le 2(a-b)\equiv d_\text{cl}$. On the other hand, the $T=0$ correlation functions behave as
%\begin{align}
%C &= \left\langle \partial_{i_1} \dots \partial_{i_{b+1}} \phi(x) \partial_{j_1} \dots \partial_{j_{b+1}} \phi(0) \right\rangle \nn
%&\sim \int \frac{d\omega\, d^dk\, k^{2({b+1})}}{\omega^2 + k^{2(a+1)}}\nn
%&\sim \int \frac{d^dk\, k^{2(b+1)}}{k^{a+1}},
%\end{align}
%so the quantum critical dimension is $d_\text{q} = a - 2b - 1$. We recover the maximal generalized Mermin-Wagner argument if we let $b=-1$ denote the trivial group ($b=0$ denotes the monopole group). Furthermore, the critical dimensions again satisfy $d_\cl = d_\text{q} + z$.
%
%
%\subsection{General formulation of Mermin-Wagner}
%% The kinetic term should take the form
%% \begin{align}
%% K[\phi(x)] &= (\partial_{i_{m+1}}\dots \partial_{i_n} \phi_{i_1 \dots i_m})^2,
%% \end{align}
%% which is invariant under shifts of the form of Eq.~\ref{eqn:partial}. The dispersion for the Goldstone modes is $K_k \sim k^{2(n-m)} = k^{2(a-b)}$. Following the correlation function calculation above, we find a classical critical dimension of $d = 2(a-b)$ and a quantum critical dimension of $d = a-b$. 
%
%\subsection{\note{Different phases for the same subgroup}} \label{sub:phases}
%
%The previous subsection shows that when the kinetic term in the effective action is $K\sim (\partial_i\partial_j\phi)^2$, there are intermediate dimensions ($d=2$ for $T>0$ or $d=1,2$ for $T=0$) where the symmetry breaking pattern preserves the monopole group. For smaller dimensions there is no symmetry breaking, while for larger dimensions (with this effective action) the group is fully broken.
%We can also write down theories where the kinetic term will be $(\partial_i\theta_j)^2$. The latter type of theory will preserve the monopole group in any large enough dimension. Since the latter theory is quadratic, it behaves like an ordinary theory, with a MW theorem prohibiting symmetry breaking for $d\le 2$ at $T>0$ or $d\le 1$ at zero temperature.
%
%We can organize these different theories that have the same SSB structure by their compressibilities. Since we are considering theories that conserve the dipole group, there will be dipole charges and monopole charges. Call the compressibility for the monopole charges $\mu_0$ and the compressibility for the dipole charges $\mu_1$. In any number of dimensions, there will be a Mott-insulator-like phase where $\mu_0=\mu_1  = 0$. The kinetic term $(\partial_i\theta_j)^2$ assumes $\mu_0=0$ and $\mu_1>0$, while the term $(\partial_i \partial_j \phi)^2$ assumes both compressibilies are nonzero.
%
%The list of compressibilies only tells us which fields are available for writing down an effective action. As in the previous subsection, which operators develop long-range order depends on the dimension. The calculations are analogous to previous correlation functions, and the results are as follows. At nonzero temperature, the order $b$ of the preserved subgroup $\mmax{b}$ is 
%\begin{align}
%\begin{tabular}{l|r|r|r}
%     & $d=1,2$ & $d=3,4$, & $d>4$ \\
%    $(\partial_i\partial_j\phi)^2$ &  1 &  0 & -1 \\
%    $(\partial_i\theta_j)^2$       &  1 &  0 &  0 \\
%    Mott                           &  1 &  1 &  1
%\end{tabular}
%\end{align}
%where we are, as always, denoting the trivial group by $b=-1$. Because the correlation calculations do not have any $\omega$ integral, the critical dimensions for a given symmetry-breaking pattern are the same in different theories.

%The same is not true at $T=0$. Similar correlation function calculations show that the symmetry-breaking patterns are 
%\begin{align}
%\begin{tabular}{l|r|r|r}
%     & $d=1$ & $d=2$, & $d>2$ \\
%    $(\partial_i\partial_j\phi)^2$ &  0 &  0 & -1 \\
%    $(\partial_i\theta_j)^2$       &  1 &  0 &  0 \\
%    Mott                           &  1 &  1 &  1
%\end{tabular} \label{eqn:zerophases}
%\end{align}
%The interesting difference from finite temperature is that the theory with a $\phi$ field can preserve $\mmax{0}$ at $d=1$, while the theory with a $\theta_j$ field cannot.
%Both the $T>0$ and the $T=0$ table can easily be generalized to higher values of $a$, the order of the preserved multipole group.

%Whenever there are distinct theories with the same symmetry-breaking pattern, some of the theories have operators with quasi-long-range order. \todo{QLRO vs SRO} For example, the $(\partial_i\theta_j)^2$  theory at $d=1$ acts like a Luttinger liquid of dipoles. A full analysis of quasi-long-range ordered fields is beyond the scope of this paper. Furthermore, we have only considered free fixed points. There certainly may be interacting theories with distinct phases and symmetry-breaking patterns~\cite{TonerRadzihovsky}. \ethan{how do you have interacting goldstones for an abelian symmetry?}

% ------------------------

% Once we have identified a symmetry group $G$ and its preserved subgroup $H$ we still have not fully specified a phase of matter. As an example, consider the previous case of partially breaking the maximal dipole group $\mathcal{M}^1_\max$ to the monopole group $\mathcal{M}^0_\max$. There is one theory with Lagrangian
% \begin{align}
% \mathcal{L}_2 = (\partial_t \phi)^2 + \sum_{ij} (\partial_i \partial_j \phi)^2,
% \end{align}
% which has critical dynamical exponent $z=2$. There is another theory with Lagrangian
% \begin{align}
% \mathcal{L}_1 &= (\partial_t \varphi_i)^2  + \sum_{ij} (\partial_i \varphi_j)^2,
% \end{align}
% and $z=1$. The two theories differ in that, if $\phi$ transforms as $\phi \goesto \phi + a_ix^i + b$, then $\varphi_i$ transforms as $\varphi_i \goesto \varphi_i + a_i$

% \charlie{Does this mean the phase transition (if it exists) has to be topological,  \ethan{in what sense?} in the sense that the symmetry group and preserved subgroups are the same/ beyond Landau?}

% For both theories, the classical critical dimension is $d_\cl = 2$. However, they have different quantum critical dimensions, $d_{\q, 1} = 1$ and $d_{\q,2}=0$. Both theories obey $d_\cl = d_\q + z$. Of course, for $d>2$ the $z=2$ theory develops long-range order for $e^{i\phi(x)}$, breaking the monopole symmetry in addition to the dipole symmetry.

% We can now generalize to arbitrary $a$ and $b$. To do so, we define a field $\phi_{i_1\dots i_k}$ which transforms under $\mathcal{M}^a_\max$ as 
% \begin{align}
% \delta \phi_{i_1\dots i_k} = \lambda_{I_c} \partial_{i_1} \dots \partial_{i_k} P_c^{I_c}.
% \end{align}
% The kinetic term for this field will be
% \begin{align}
% K \sim (\partial_{i_{k+1}} \dots \partial_{i_{a+1}} \phi_{i_1\dots i_k})^2, \label{eqn:atok}
% \end{align}
% so that $z = 1+a-k$. Although $\phi_{i_1\dots i_k}$ is invariant under $\mathcal{M}^k_\max$, that field may not be able to develop long-range order. The fields that do develop LRO may be invariant under larger groups. In particular, $\partial_{i_{k+1}} \dots \partial_{i_{b+1}} \phi_{i_1\dots i_k} $ is invariant under $\mathcal{M}^b_\max$

% At finite temperature, the field $\phi_{i_1\dots i_k}$ develops long-range order when $d>2(a-k)$, so that the preserved subgroup is $\mmax{k}$. When $2(a-b)<d\le 2(a-b+1)$, the preserved subgroup is $\mmax{b}$. The upshot is that there are $N^d_{ab}$ to spontaneously break the symmetry from $\mmax{a}$ to $\mmax{b}$ in $d$ spacial dimensions. The function is
% \begin{align}
% N^d_{ab} = \begin{cases} 0, \quad d\le 2(a-b) \\ 2 \\ 
% \end{cases}
% \end{align}

% All of these theories will have $d_\cl = 2(a-b)$ and $d_\q = 2(a-b) - z$.

% In addition to having different critical exponents $z$, these different theories with the same symmetry breaking patterns have different numbers of Goldstone modes. For the dipole examples considered above, the $z=1$ theory has $d$ massless modes (the components of $\varphi_i$) while the $z=2$ theory only has a single massless mode, $\phi$. \charlie{What else should we say about this?}

% \todo{How should we address the question of stability?}
% \ethan{I think my comments about stability are actually not relevant here --- they only imply that the transition between the ordered and disordered phases cannot be studied using by doing a perturbative analysis of the effects of vortex operators in the ordered phase}

@@

The term $H_\text{int}$ is allowed by the symmetry because it simultaneously removes an edge boson and creates a dipole of site bosons. Note that when the edge bosons are condensed, $H_\text{int}$ gives the site bosons an effective single particle hopping: thus condensation of single bosons can either proceed directly (by making $t_b$ large) or indirectly (by first condensing dipoles, and making the effective single-particle hopping term large). 
%and hence an effective $k^2$ dispersion, which would otherwise be disallowed by symmetry. \charlie{What would the effective dispersion actually be? Does it make sense to say that it is $k^2$ at short distances and $k^4$ at long distance?}

%\charlie{Relevance of cosine terms and stability to crystal phase?}
%\ethan{at large enough hopping strength we will generically just get a condensate and don't have to worry about this stuff}

@@

%The phases of this model will be those in Eqn.~\ref{eqn:zerophases}. The values of $t_b$ and $t_d$ will determine which fields are available for the IR action. When the site bosons are condensed the $\phi$ field exists but might not be long-range ordered. Similarly when the edge bosons are condensed the $\theta_j$ field exists but might not be long-range ordered. For $d<2$ any symmetry-breaking pattern is possible. For $0<d\le 2$ the symmetry can not be completely broken, but the dipole part can be broken.

%At sufficiently high dimension, this system will support three phases, which may be loosely described as the Mott insulator, the monopole superfluid where site bosons are condensed and the dipole superfluid where edge bosons (but not site bosons) are condensed. 
%Note that the dipole superfluid does preserve lattice rotations.
%In the monopole superfluid phase, the effective Hamiltonian contains bare creation and annihilation terms for the edge bosons, which automatically breaks number conservation of the edge bosons, and effectively induces a dipole superfluid as well. We can identify these phases with symmetry breaking patterns of the multipole group. In the trivial phase the group is unbroken. In the dipole superfluid, the group is broken to its monopole subgroup. The monopole superfluid then corresponds to the fully broken symmetry.

% In the absence of disorder, neither symmetry breaking pattern can occur for $d\le 1$. There may still be interesting phases, such as a Luttinger liquid.
% For $d>2$ dipole and monopole symmetry breaking may occur, so all three phases are intact. Furthermore, transitions between any two phases are allowed.
% This leaves the intermediate case of $d=2$, in which all three phases can exist, but there can be no transitions between any symmetry-unbroken phase and the monopole superfluid. Instead, there must be an intermediate phase with dipole but not monopole superfluidity. Representative single-parameter phase diagrams are shown in Fig.~\ref{fig:phases}.

% \begin{figure}
% \centering

% \begin{tikzpicture}[scale = .9]
% \draw[<->] (0,0) node[left]{(a)}  -- (8,0) node[right]{$s$};
% \draw (4,-.1) -- (4,.1);

% \node [align=center] (A) at (2, -0.5) {unbroken};
% \node [align=center] (C) at (6, -0.5) {fully\\broken};

% \end{tikzpicture}
% \begin{tikzpicture}[scale = .9]

% \draw[<->] (0,0) node[left]{(b)}  -- (8,0) node[right]{$s$};
% \draw (2.667,-.1) -- (2.667,.1);
% \draw (5.333,-.1) -- (5.333,.1);

% \node [align=center] (A) at (1.333, -0.5) {unbroken};
% \node [align=center] (B) at (4, -0.5) {dipole\\broken};
% \node [align=center] (C) at (6.667, -0.5) {fully\\broken};
% \end{tikzpicture}

% \caption{\todo{Remove figure} \todo{New figure with different dipole-breaking phases?}(a) For  $d>2$, there can be direct transitions between symmetry-unbroken and the fully-broken phase. Here $s$ is intended to represent a parameter along an arbitrary line in the phase diagram that crosses this transition. (b) At $d=2$, although the symmetry can be fully broken, the fully broken phase cannot touch the unbroken phase. Any line through the phase diagram that connects the two phases must pass though the dipole-broken phase. These phase diagrams are for translation invariant systems at zero temperature.}
% \label{fig:phases}
% \end{figure}

@@

%At first, we will set $h_d=0$ but turn on $h_b$. This corresponds to the main considerations from Sec.~\ref{sec:disord}, where the disorder couples to the order parameter that fully breaks the symmetry. This raises the critical dimension for the long-range order of the site bosons. The edge bosons may still enter a superfluid phase for $d>1$, but the site bosons may not condense with long range phase order until $d>4$. There cannot be a direct transition from the insulator to the monopole superfluid until $d>8$.  A phase akin to a Bose glass \cite{Fisheretal} of monopoles, where site bosons condense but without long-range order, can still exist in $1\le d<4$.

%Now, let $h_d\ne 0$ while keeping $h_b=0$. The Imry-Ma argument for the edge bosons says that they should not be able to enter a superfluid phase when $d\le 4$. Once again, we can enter a Bose glass phase where $\langle d_{j, j+\mu} \rangle \ne 0$. In this case, the site bosons acquire a random single particle {\it hopping}. This may be expected to give rise to a Bose glass phase also for the site bosons. This implies that the monopoles cannot even undergo mediated condensation to superfluidity for $d\le 4$. For $d>4$, the dipoles may enter a superfluid phase, and the monopoles can enter a superfluid phase without mediation.

%It is difficult to justify including including one disorder field but not the other. The standard reason to forbid disorder is by imposing translation invariance, but if we are already including one disorder field we are necessarily breaking translation symmetry. 
For $h_b,h_d\ne 0$, we can fully rely on the Imry-Ma argument. No symmetry breaking can occur for $d\le 4$, the dipole part of the symmetry may be broken for $4<d\le 8$, and any symmetry-breaking phase can occur for $d>8$.

>>>>

Reference~\cite{Gromov2019} includes discussion of some multipole groups; we will review a few here. The simplest case is the maximal multipole group $\mathcal{M}^a_\text{max}$, which includes all shifts by polynomials of degree $a$ or less. Individual monomials can be written as
\begin{align}
P_c^{\gamma} = \mu^{\gamma}_{I_c}x^{I_c}, \label{eqn:basis}
\end{align}
where $\gamma$ is an index that runs over monomials of degree $c$ and $c\le a$. The composite index $I_c = \{i_1,\dots,i_c\}$  expands as $x^{I_c}=x^{i_1}\cdots x^{i_c}$, and similarly for $\partial_{I_c}$. Each  matrix  $\mu^{\gamma}$ is fully symmetric on its $I_c$ indices. This group also includes all translations and rotations.

An example of a multipole group that contains all translations and rotations but is \emph{not} the maximal multipole group is the group generated by shifts of the form
\begin{align}
\phi \goesto \phi + \lambda_0 P_0^0 + \lambda_{i} P^{i}_1 + \lambda_{\beta} P^{\beta}_2,
\end{align}
where the degree-0 polynomial is $P_0^0=1$. The other polynomials are
\begin{align}
P_1^{i} = x^i,\quad \quad P_2^{\beta} &= \mu^{\beta}_{ij} x^i x^j,
\end{align}
where $\mu^{\beta}$ are a basis for the rank-2 traceless symmetric $d\times d$ matrices. Let us call this group $\mathcal{M}^2_{\text{sym}}$. Recall that the maximal quadrupole group $\mathcal{M}^2_{\text{max}}$ is already only built from symmetric matrices $\mu$. The tracelessness condition thus only removes one polynomial from the set.

@@

We can now consider building IR field theories for phases in which a multipolar symmetry is completely spontaneously broken, with no residual unbroken subgroup. Reference~\cite{Gromov2019} describes the process in detail, so in the following we will be somewhat succinct. 
We will continue to write things down in terms of the compact scalar $\phi$, which (at in the cases for which SSB is not preempted by strong fluctuations) is to be viewed as the Goldstone for the sponatneously broken symmetry.

Since we are only interested in the low-energy physics of the putative symmetry-broken phase, constructing an appropriate IR field theory amounts to nothing more than constructing an appropriate kinetic term for $\phi$. 
To find a kinetic term invariant under a given multipole group $\mathcal{M}$, we need to find operators $D$ built out of spatial derivatives that annihilate all the polynomials in $\mathcal{M}$. If $a_\text{max}$ is the highest degree of the polynomials in $\mathcal{M}$, the simplest derivative operators which generically do the job are of the form 
\begin{align}
D = q^{I_{a_{\rm max}+1}} \p_{I_{a_{\rm max}+1}},
\end{align}
for some symmetric tensor $q^{I_{a_{\rm max}+1}}$. Although it is not generically possible to do so~\cite{Gromov2019}, we can sometimes find a set of $D_\alpha$ ($\alpha$ is an abstract index) with $s\le a_{\text{max}}$, where $a_\text{max}$ is the highest degree of the polynomials. In this case, the effective field theory will be invariant under some non-maximal multipole symmetry.

Writing the invariant derivative operators as $D_\alpha$, the most general kinetic term is ~\cite{Gromov2019} 
\begin{align}
K[\phi(x)] &= g_{\alpha\beta} (D_\alpha \phi) (D_\beta \phi)
\end{align}
for some symmetric tensor $g_{\alpha\beta}$.
We will often write the Fourier transform of the kinetic term as $\phi_{-k} K_k \phi_k$.  Requiring some spatial symmetries restricts the choices of $g_{\alpha \beta}$. For the maximal multipole group, enforcing all rotation symmetries results in the kinetic term
\begin{align}
K_a[\phi(x)] &= (\partial_{I_{a+1}} \phi) (\partial_{I_{a+1}} \phi),
\end{align}
which is the kinetic terms studied in Ref.~\cite{Griffin2015}.

@@

Consider first a clean system at $T>0$ with a spontaneously broken maximal multipole symmetry of degree $a$.
Then the kinetic term is proportional to 
\begin{align}
K[\phi(x)] = (\partial_{I_{a+1}} \phi)^2. \label{eqn:adisp}
\end{align} 
Now consider nucleating domains of linear size $L$ in which the expectation value of the order parameter --- namely $\langle e^{i\phi}\rangle$ --- varies by an amount of order 1. Since the order parameter is continuous there will be a thick domain ``wall" over which the order parameter changes gradually from its value in the background phase to its value in the domain. The thickness will be of order $L$.
Equation~\ref{eqn:adisp} tells us the  energy cost of the field changing its value will be minimized if it varies as a polynomial with degree $a+1$. In that case the energy density will scale as $K_k \sim k^{2(a+1)} \sim L^{-2(a+1)}$. Integrating that energy density over a region of size $L$ gives a total energy cost to nucleating domains of $L^{d-2(a+1)}$.

@@

We can also consider an anisotropic multipole group. 
Let the conserved charges be the monopole moment and $d_2$ components of the dipole moment. Then the dispersion will be $k^4 + p^2$, where $k$ has $d_2$ components, $p$ has $d_1$ components, and $d = d_1+d_2$. At nonzero temperature, the correlation function is
\begin{align}
C &\sim \int \frac{d^{d_2} k \, d^{d_1} p}{k^4 + p^2}\nn
&\sim \int x^{d_2/2 -1} p^{d_1-1} \frac{dx\, dp}{x^2 +p^2} \nn
&\sim \int q^{d_2/2 + d_1 - 2}\frac{dq}{q},
\end{align}
where $q^2 = x^2 + p^2 = k^4 + p^2$. 
If $d_1=2$ the symmetry can be broken for any $d_2>0$, and if $d_1=1$ then $d_2=2$  is critical. Of course, if $d_1=0$ we have an isotropic quartic dispersion and $d=d_2=4$ is critical.

@@

The $T=0$ correlation function behaves as
\begin{align}
C &\sim \int \frac{d\omega d^{d_2} k \, d^{d_1} p}{\omega^2 + k^4 + p^2}\nn
&\sim \int k^{d_2 - 2} p^{ (d_1-1)} \frac{k\,dk\, dp}{\sqrt{k^4 + p^2}} \nn
&\sim \int x^{d_2/2-1} p^{d_1-1} \frac{dx\, dp}{\sqrt{x^2 + p^2}} \nn
&\sim \int q^{d_2/2 -1} q^{d_1 -1} dq,
\end{align}
where again $q^2 = x^2 + p^2 = k^4 + p^2$. This diverges when $d_2/2 + d_1 \le 1$.

The previous analysis suggests a procedure for finding the critical dimension in clean anisotropic systems for breaking the multipole symmetry to the trivial group. First, sort each dimension by the degree of its dispersion relation, so that there are $d_n$ dimensions with dispersion $k^{2n}$. Then define the effective dimension as 
\begin{align}
d_\eff = \sum_n \frac{d_n}{n}.
\end{align}
For $T>0$, we then conclude that SSB cannot occur if $d_\eff\le 2$, while at $T=0$ the critical dimension is $d_\eff = 1$.

@@

Let us define the compressibilities $\kappa_c$ by 
\be \label{kappadef} \kappa_c^{I_c} = \frac{d\langle n_c^{I_c}\rangle}{d\mu_c^{I_c}},\ee 
where $n_c$ is the number density of $c$-pole charges, $\mu_c$ is a conjugate chemical potential, and $I_c = \{i_1,\dots,i_c\}$ is a composite index as before.

Let $b\leq a$ denote the smallest degree of multipole charge with nonzero compressibility, i.e. consider a situation in which $\kappa_d>0$ for all $b\leq d\leq a$, while $\kappa_d=0$ for all $d<b$. Then the minimal IR Lagrangian for the symmetry broken phase looks schematically like 
\be \label{generall} \mathcal{L} = \kappa_b (\partial_\tau \phi_{I_b})^2 + g_{J_{a-b+1}}(\partial_{J_{a-b+1}}\phi_{I_b})^2,\ee 
where 
$\p_{J_{a-b+1}} \equiv \p_{j_1} \cdots \p_{j_{a-b+1}}$, and where $\phi_{I_b}$ is a field that shifts by a polynomial of degree $a-b$ under the action of $\mmax a$. The compressibility appears as the coefficient of the time derivative term since $\p_\tau \phi_{I_b}$ is the momentum conjugate to $\phi_{I_b}$ in the quantum theory, and hence correlation functions of fluctuations in $n_b$ are determined by those of $\p_\tau \phi_{I_b}$.

@@

However, a nonzero $\kappa_b$ does not {\it necessarily} imply that $\mmax b$ is spontaneously broken. $\kappa_b\neq 0$ means that the system possesses gapless modes created by $n_b$, but it is possible for fluctuations to be strong enough so that these modes cannot be associated with the Goldstone modes of a spontaneously broken $\mmax b$. Thus the theory \eqref{generall} does not necessarily describe a phase where $\mmax b$ is spontaneously broken, and the exact pattern of symmetry breaking depends on the spatial dimension and whether or not $T$ is nonzero. For example, we can again consider the theory in \eqref{singlescalar} at $d=3, T>0$: even though the charge compressibility $\kappa_0$ (the coefficient of $(\p_\tau\phi)^2$ in \eqref{singlescalar}) is nonzero, $\phi$ fluctuates strongly enough that the monopole symmetry is not spontaneously broken.

@@

@@

The term $H_\text{int}$ is allowed by the symmetry because it simultaneously removes an edge boson and creates a dipole of site bosons. Note that when the edge bosons are condensed, $H_\text{int}$ gives the site bosons an effective single particle hopping: thus condensation of single bosons can either proceed directly (by making $t_b$ large) or indirectly (by first condensing dipoles, and making the effective single-particle hopping term large).

@@

@@

For $h_b,h_d\ne 0$, we can fully rely on the Imry-Ma argument. No symmetry breaking can occur for $d\le 4$, the dipole part of the symmetry may be broken for $4<d\le 8$, and any symmetry-breaking phase can occur for $d>8$.
